---
title: "Documenting R Code"
author: "Kelly Sovacool"
date: "5/21/2019"
output: 
    ioslides_presentation:
        widescreen: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Why document your code?

- Reproducibility
- FAIR data principles: findable, accessible, **interopable**, **reusable**
- Make it usable for other scientists -- including future you!

## Ten simple rules for documenting scientific software

### Excerpted from [Lee et al. 2018](https://doi.org/10.1371/journal.pcbi.1006561):

- Write comments as you code.
- Include a README file with basic information.
- Version control your documentation.
- Use automated documentation tools.
- Write error messages that provide solutions or point to your documentation.

## What not to document

- "Document Design and Purpose, Not Mechanics" - [Wilson et al. 2014](Wilson G, Aruliah DA, Brown CT, Chue Hong NP, Davis M, Guy RT, et al. (2014) Best Practices for Scientific Computing. PLoS Biol 12(1): e1001745. https://doi.org/10.1371/journal.pbio.1001745)
- In other words, "Why, not how".
- Don't do this:
    ```{r}
    i <- i + 1  # increment `i` by 1
    ```
- Try to write code that is self-documenting.
    - Use descriptive variable names. If you have to write a comment to describe your variable, you probably need to rename it. Examples:
        - `counter` instead of `i`
        - `patient_metadata` instead of `df`
        - `data_cleaned` instead of `df2`
    
## What to document

- No code is fully self-documenting. 
- You should document:
    - packages
    - datasets
    - functions
    - classes
    - any tricky lines of code

## How to document

### The bare minimum:

Include a comment at the top of your R script to briefly describe what it does at a high level. 

```{r}
# Generate plots from mothur sensspec files for comparing clustering algorithms.
library(ggplot2)
```

## How (specifics for R)

### Best practices:

- For any project with R scripts:
    - Write comments alongside your code with `roxygen2` syntax for `man/` files.
    - Make your project a package with `usethis` & `devtools`.
- For packages you release into the wild:
    - Write a vignette with `R Markdown`.
    - Create a website with `pkgdown`.

## `roxygen2` syntax

[Document functions](http://r-pkgs.had.co.nz/man.html) in `R/*.R` files
```{r}
#' Add together two numbers.
#' 
#' @param x A number.
#' @param y A number.
#' @return The sum of \code{x} and \code{y}.
#' @examples
#' add(1, 1)
#' add(10, 1)
add <- function(x, y) {
   x + y
}
```

## `roxygen2` syntax

[Document datasets](http://r-pkgs.had.co.nz/data.html) in `R/data.R`
```{r}
#' Prices of 50,000 round cut diamonds.
#'
#' A dataset containing the prices and other attributes of almost 54,000
#' diamonds.
#'
#' @format A data frame with 53940 rows and 10 variables:
#' \describe{
#'   \item{price}{price, in US dollars}
#'   \item{carat}{weight of the diamond, in carats}
#'   ...
#' }
#' @source \url{http://www.diamondse.info/}
"diamonds"
```

## `roxygen2` syntax

[Document the package](http://r-pkgs.had.co.nz/man.html) in `R/package_name.R`
```{r}
#' foo: A package for computating the notorious bar statistic.
#'
#' The foo package provides three categories of important functions:
#' foo, bar and baz.
#' 
#' @section Foo functions:
#' The foo functions ...
#'
#' @docType package
#' @name foo
NULL
```

## Activity

Let's document code from the [Riffomonas minimalR](http://www.riffomonas.org/minimalR/06_line_plots.html) tutorial!

1. the package: edit [R/minimalR.R](R/minimalR.R)
1. datasets: edit [R/data.R](R/data.R)
    - `baxter_metadata` 
    (raw data is in `inst/extdata` & processed data is in `data/`)
1. functions: edit [R/baxter.R](R/baxter.R)
    - `get_metadata` & `get_bmi`
    - `get_bmi_category` & `is_obese`

## Activity

1. Clone this repo
    ```{bash}
    git clone https://github.com/SchlossLab/documenting-R
    ```
    or if you previously cloned it, pull new commits:
    ```{bash}
    cd path/to/documenting-R/
    git pull
    ```
1. Checkout a new branch
    ```{bash}
    git checkout -b descriptive-branch-name
    ```

## Activity

1. After modifying your part, commit your changes
    ```{bash}
    git add . ; git commit -m "descriptive commit message"
    ```
1. Push your changes
    ```{bash}
    git push -u origin descriptive-branch-name
    ```
1. Open a pull request on GitHub to merge your branch into master. 
    - Mention your issue number(s) in the description like `Resolves #issue_number`.

## R Packages 

Setup an R package
```{r}
library(usethis)
library(devtools)
create_package(file.path(getwd()))
```

Specify dependencies
```{r}
use_package("dplyr")
use_package("readxl")
```

Compile documentation
```{r}
devtools::document()
```

## Additional Reading

- [R Packages](http://r-pkgs.had.co.nz/man.html) - Hadley Wickham
- [pkgdown](https://pkgdown.r-lib.org/): build a website for your R package
- [Mastering Software Development in R](https://bookdown.org/rdpeng/RProgDA/documentation.html) - Roger Peng, Sean Kross, & Brooke Anderson

